A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-30 paper.tex

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} ARGs inferred by
(A) \kwarg, (B) \argweaver, (C) \tsinfer,
and (D) \relate\
for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
locus~\citep{kreitman1983nucleotide}. 
The \argweaver\ example
is a sample from the inferred posterior distribution. In each ARG,
colours are assigned to edges based on the time of the edge's child node
(lighter: older; darker: younger).
Position on the X and Y axes for the graph visualization is arbitrary.
% Note that the position of points on the
% Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
% algorithm, rather than as a direct measure of time.
Line width corresponds to the genomic span of the edge. Node colour
refers to the number of parents of the node.
The fill colour intensity of a node is determined by the genomic span over which
it has two or more children divided by the span where it has any
children (the ``coalescent span''). 
The genomic intervals associated with edges are shown below
each ARG, where each row is an edge
and the X axis reflects genomic position. 
% Edges
% are ordered by the age of their child node, from oldest to youngest
% moving down the page, with colours matching the edge colours in the ARG.
Vertical lines indicate breakpoints between local trees.
}
\end{figure}

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} varying levels of precision
about recombination in an ARG. 
It is also possible, of course, to \emph{infer} varying levels of precision about 
recombination.
To illustrate this, and to demonstrate the structural 
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. 
Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees, and hence lie somewhere 
in between a sequence of unrelated trees on one extreme and a 
completely precise event ARG on the other, in terms of the precision
inferred about recombination.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is not as simple as it may seem.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.

Firstly, the internal nodes in the local trees must be labelled
(i.e.\ marked with the identity of the corresponding ARG node).
If the nodes are not labelled, we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.
% JK: Does this need to be said?
% (One may use the time of a node as a proxy for identity,
% but it is not reliable as we can have multiple nodes at the
% same time, e.g. with time discretisation.)

The precision of the information about recombination that
can be represented in the trees depends the ARG itself.
Secondly, the local trees must contain ``unary'' nodes.
[Some discussion].
With these two conditions met, there is a direct correspondance
between a gARG and the trees.

These properties have interesting consequences when we
consider the inference strategy of esimating local trees
independenty and subsequently stitching them together
into an ARG. Two recent methods take this approach,
\relate\ and
\texttt{espalier}~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local
trees into an ARG. [Relate does something fairly lightweight,
but by default imprecise.]
Espalier, on the other hand, is specifically in interested
in the recombination events and therefore attempts to infer
the precise ``subtree prune-and-regraft'' (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
using heuristics~\citep{rasmussen2022espalier}.
Inferring the SPRs between
leaf-labelled trees is NP-hard in
general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

@@

A necessary first step towards this goal is some terminological
clarity. As discussed in Appendix XXX, the term ARG has several
subtly different interpretations depending on context. The
trend to decouple the ARG from its origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as being any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.

The flexiblity of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which, because of its origins in an idealised
mathematical model, is more limited in what can be described.
The necessity of complete precision under the eARG encoding has been
conflated with the idea of a ``true'' or ``full'' ARG.
[A discussion of how realistic inferring a completely precise ARG is]
[Do downstream applications use full ARGs? usually no]

>>>>

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference results for
(A) \kwarg, (B) \argweaver, (C) \tsinfer,
and (D) \relate\
for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
locus~\citep{kreitman1983nucleotide}. 
The \argweaver\ example
is a sample from the inferred posterior distribution.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
Line width corresponds to the genomic span of the edge.
Position on the X and Y axes for the graph visualization is arbitrary.
 Node colour
refers to the number of parents of the node.
The fill colour intensity of a node is determined by the genomic span over which
it has two or more children divided by the span where it has any
children (the ``coalescent span''). 
The genomic intervals associated with edges are shown below
each ARG, where each row is an edge
and the X axis reflects genomic position. 
Vertical lines indicate breakpoints between local trees.
}
\end{figure}

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} varying levels of precision
about recombination in an ARG. 
It is also possible, of course, to \emph{infer} varying levels of precision about 
recombination.
To illustrate this, and to demonstrate the structural 
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. 
Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees, and hence,
in terms of the level of detail inferred about recombination,
lie somewhere 
in between a sequence of unrelated trees on one extreme and a 
completely precise event ARG on the other.

@@

% \section*{Equivalance of ARGs and the set of local trees}
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is not as simple as it may seem.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.

% Firstly, the internal nodes in the local trees must be labelled
% (i.e.\ marked with the identity of the corresponding ARG node).
% If the nodes are not labelled, we must infer their identity
% from the tree topology, which is both
% expensive to compute and imprecise. For example
% we could infer that node \textsf{k} is shared in the second
% and third trees of in Fig.~\ref{fig-simplification}D
% because it subtends identical subtrees in both,
% but it would not be straightforward to infer that it
% is also the root of the first tree.
% % JK: Does this need to be said?
% % (One may use the time of a node as a proxy for identity,
% % but it is not reliable as we can have multiple nodes at the
% % same time, e.g. with time discretisation.)

% The precision of the information about recombination that
% can be represented in the trees depends the ARG itself.
% Secondly, the local trees must contain ``unary'' nodes.
% [Some discussion].
% With these two conditions met, there is a direct correspondance
% between a gARG and the trees.

% These properties have interesting consequences when we
% consider the inference strategy of esimating local trees
% independenty and subsequently stitching them together
% into an ARG. Two recent methods take this approach,
% \relate\ and
% \texttt{espalier}~\citep{rasmussen2022espalier},
% and have different strategies for reconciling the local
% trees into an ARG. [Relate does something fairly lightweight,
% but by default imprecise.]
% Espalier, on the other hand, is specifically in interested
% in the recombination events and therefore attempts to infer
% the precise ``subtree prune-and-regraft'' (SPR)
% operations~\citep{hein1990reconstructing,song2003on,song2006properties}
% using heuristics~\citep{rasmussen2022espalier}.
% Inferring the SPRs between
% leaf-labelled trees is NP-hard in
% general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

@@

A necessary first step towards this goal is some terminological
clarity. As discussed in Appendix XXX, the term Ancestral Recombination
Graph has several
subtly different interpretations depending on context. The
trend to decouple the ARG from its origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as being any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient. 
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which, because of its origins in an idealised
mathematical model, is more limited in what can be described.

The necessity of complete precision about all recombination
events under the eARG encoding has been
conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
For example, \citet{brandt2021evaluation} define a ``full'' ARG as
``a structure that encodes all
coalescence and recombination events''. While there are certainly cases
when inferring such fully precise ARGs is useful, for example when 
studying the details of a small number of recombination
events~\citep{rasmussen2022espalier}, it is worth considering
whether sufficient information exists to support such inferences
and indeed whether this precision is required (or indeed useful
for) downstream applications.

[Discuss information about recomb]
\citet{shipilina2022origin}

% % Do downstream applications actually make use of such full ARGs?
Downstream applications using estimated ARGs to date rarely
make use of such complete recombination information.
The \relate\ selection statistic~\citep{speidel2019method}
computes a $p$-value for selection [FIXME] for each
local tree independently by comparing the rates of coalescence
at various time-points. For cases like strong positive selection,
where we expect there to be a clear signal in the local tree,
such tree-by-tree methods can be simple and powerful.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in a ``full'' ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
for ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
[MORE applications]

% A major strength of this approach is that it can be used on
% any gARG, ranging from estimates containing relatively little
% information about recombination (e.g., as estimated by \relate)
% to ``full'' ARGs containing complete recombination information
% (e.g., as estimated by \kwarg), and will fully utilise
% any between-tree correlation information that is present.
